---
title: "STATS331Final Q"
subtitle: "Final Project"
author: "Ricky Murillo, Felix tran, Nathan Madlansacay, Jacob Hinshaw"
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
    number-sections: true
    code-fold: true
editor: visual
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

```{r setup}
#| include: false
library(tidyverse)
library(ggplot2)
set.seed(1)
```

## Dataset

For this project, we are working with two datasets, murders/100k people and armed forces personnel as a percent of the labor force. Both of these datasets give their respective variable per country and in every available year, which ranges from 1950 to 2020. The murder data is sourced from the Global Burden of Disease Study 2017 (<https://ghdx.healthdata.org/gbd-2017>) and the armed forces data is sourced from the International Institute for Strategic Studies.

```{r}
Soldierdata <- read.csv("armed_forces_personnel_percent_of_labor_force.csv")
Murderdata <- read.csv("murder_per_100000_people.csv")
```

### Hypothesis

We expect a negative relationship, where countries with more military personnel tend to have lower murder rates due to stronger security and law enforcement.

## Cleaning Data

```{r}
colnames(Soldierdata) <- str_replace(colnames(Soldierdata), "^X", "")
colnames(Murderdata) <- str_replace(colnames(Murderdata), "^X", "")

Soldierdata$country <- str_trim(Soldierdata$country)
Murderdata$country <- str_trim(Murderdata$country)

Soldierdata[is.na(Soldierdata)] <- 0
Murderdata[is.na(Murderdata)] <- 0

Soldierdata_long <- pivot_longer(Soldierdata, cols = -country, names_to = "year", values_to = "soldiers")
Murderdata_long <- pivot_longer(Murderdata, cols = -country, names_to = "year", values_to = "murder_rate")

FinalData <- inner_join(Soldierdata_long, Murderdata_long, by = c("country", "year")) |>
  filter(soldiers != 0,
         murder_rate != 0)



FinalData |>
  slice_head(n = 5) 
```

After done cleaning the data, the dataframe spans from the years 1990 to 2016. We did these years because the other years did not include any values and both data sets had to include an observation. Missing values were addressed by removing them from our point of focus, ensuring that all available data points were retained. The country was only included if we had data from our two test variables from the year 1990 to 2016. The solider data had 172 observations and crime rate data had 118 observations. After doing our data cleaning we had a final dataset of 85 country observations. An observation is consisted with the country, murder rate, and solider rate.

## Linear Regression

```{r}
summary_data <- FinalData |>
  group_by(country) |>
  summarize(
    avg_soldiers = mean(soldiers, na.rm = TRUE),
    avg_murder_rate = mean(murder_rate, na.rm = TRUE)) 


ggplot(data = summary_data, aes(x = avg_soldiers, y = log(avg_murder_rate))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Average Military Personnel vs. Average Murder Rate Across Countries",
       x = "Military Personnel (% of Labor Force)",
       y = "Murder Rate (per 100,000 people)")

```

The data was first aggregated so that the country observations were taking an average across the years. After, a log transformation was done the y variable of murder rate in order to create a linear regression and try to create an equal variance with our residuals.

As seen on the scatter plot, the distribution is focused towards the **bottom left** of the graph **(under x = 10 and y = 25)**. This indicates that most countries have a relatively **small percentage of military** personnel in their labor force. However, among those countries, there is a wide variation in murder rates, with some exceeding **25 per 100,000 people**.

A key pattern observed in this **regression analysis** is the **inverse relationship** between military personnel percentage and the murder rate. Countries with a **higher percentage of military personnel** tend to have **lower or negligible murder rates**. Conversely, those with **higher murder rates** generally have a **smaller percentage of military personnel** in their labor force.

Overall, the data shows a **strong skew** to the right, where **high murder rates** are concentrated in regions with a **low percentage of military personnel**, while **lower murder rates** are observed in countries with a **higher proportion of military personnel**.

```{r}
summary_data_year <- FinalData |>
  group_by(year) |>
  summarize(
    avg_soldiers = mean(soldiers, na.rm = TRUE),
    avg_murder_rate = mean(murder_rate, na.rm = TRUE))

ggplot(summary_data_year, aes(x = as.numeric(year))) +
  geom_line(aes(y = avg_murder_rate, color = "Murder Rate"), size = 1) +
  geom_line(aes(y = avg_soldiers * 2, color = "Military Personnel (Scaled)"), size = 1, linetype = "dashed") +
  scale_y_continuous(name = "Murder Rate (per 100,000 people)",
                     sec.axis = sec_axis(~./2, name = "Military Personnel (% of Labor Force)")) +
  labs(title = "Trends in Murder Rate and Military Personnel Over Time",
       x = "Year") +
  scale_color_manual(values = c("Murder Rate" = "red", "Military Personnel (Scaled)" = "blue")) +
  theme_bw()

```

## Summary(Linear Model)

This graph presents the murder rate per 100,000 people from 1990 to 2016 on a global scale, shown by the *red line*. Alongside it, the *blue dashed line* represents the **percentage of military personnel** within the labor force over the same period. The **x-axis marks** the years, while the left **y-axis** corresponds to the **murder rate**, and the right y-axis corresponds to the **percentage of military personnel**. This visualization displays how both variables have changed over time.

## Analysis(Linear Model)

The graph shows that the **murder rate** increased significantly between **1990 and 1995**, reaching its peak before steadily declining. The **percentage of military personnel within the labor force**, represented by the **blue dashed line**, also experienced a rise around the same period before gradually decreasing. This **overlapping trend** suggests a possible relationship between the two variables, though further analysis would be required to determine the nature of the connection. While the murder rate began to fall after its peak in the mid-1990s, the percentage of military personnel also saw a decline, indicating that both variables may have been influenced by broader socio-political or economic factors during this time.

```{r}
model <- lm(log(avg_murder_rate) ~ avg_soldiers, data = summary_data)
summary(model)
coefficients(model)
```

In table above is the summary statistics for our regression model. We use a linear regression to compare the effects military personal has on murder rate. We use the same transformation as used in the scatter plot above with applying a log transformation to the y variable of murder rate. From this table we get the parameter estimates and the R-squared value.

Linear Model: \$ \\hat{\\text{Murder Rate}} = 1.0924 - 0.1101 \\times \\text{Military Personnel} \$

Intercept :When military personnel is 0%, the predicted murder rate is 1.0924 per 100,000 people.

Slope: For every 1% increase in military personnel as a share of the labor force, the murder rate is predicted to decrease by 0.1101 per 100,000 people

```{r}
fitted_values <- fitted(model)
residuals <- resid(model)

variance_response <- var(summary_data$avg_murder_rate, na.rm = TRUE)
variance_fitted <- var(fitted_values, na.rm = TRUE)
variance_residuals <- var(residuals, na.rm = TRUE)

library(knitr)
library(kableExtra)

data.frame(
  "Variance in Response" = variance_response,
  "Variance in Fitted" = variance_fitted,
  "Variance in Residuals" = variance_residuals
) |>
  kable() |>
  kable_styling(full_width = FALSE)

r_squared <- variance_fitted / variance_response
r_squared
```

The R-squared value or the fitted variance divided by the response of variance is 0.0129. This shows that only 1.29% of the variation in murder rates is explained by the percentage of armed forces personnel in the labor force. This very low value suggests that military employment has little to no meaningful impact on murder rates, at least alone in the regression model. Most of the variability, 96.76%, is due to other factors that are not included in the model. Because crime rates are typically influenced by a mix of socioeconomic, demographic, and policy-related variables, relying on military employment as a predictor is unlikely to provide a strong explanatory factor.

To improve the model, we would need to add additional predictors that are more directly linked to crime rates. Factors such as unemployment, poverty levels, education, and law enforcement policies could offer greater variance into the drivers of violent crime. Given the weak variance of the current model, a broader approach that includes multiple predictors would likely provide a more accurate and comprehensive understanding of what influences murder rates.

## Simulation

```{r}

predictions <- predict(model)
residual_sd <- sigma(model)

simulated_data <- data.frame(
  avg_soldiers = summary_data$avg_soldiers,
  simulated_murder_rate = pmax(predictions + rnorm(length(predictions), mean = 0, sd = residual_sd), 0)

)

Thousand_Simulated <- map_dfr(
  .x = 1:1000,
  .f = function(i) {
    data.frame(
      avg_soldiers = summary_data$avg_soldiers, 
      simulated_murder_rate = pmax(predictions + rnorm(length(predictions), mean = 0, sd = residual_sd), 0)
    )
  }
)




ggplot() +
  geom_point(data = summary_data, aes(x = avg_soldiers, y = log(avg_murder_rate)), color = "blue", alpha = 0.5) +
  geom_point(data = simulated_data, aes(x = avg_soldiers, y = log(simulated_murder_rate)), color = "red", alpha = 0.5) +
  labs(title = "Observed vs. Simulated Data",
       x = "Military Personnel (% of Labor Force)",
       y = "Murder Rate (per 100,000 people)",
       color = "Dataset") +
  scale_color_manual(values = c("Observed" = "blue", "Simulated" = "red")) +
  theme_minimal()


```

In the model above, it shows the original observed data in blue and and simulated data in red. The simulated data has gone through the same transformation as the observed data. Military Personal is on the x-axis and murder rate is on the y-axis. The simulated and observed data both cluster at low military percentages. The observed data has more variation in murder rates. The simulated data stays mostly near zero and misses the extreme values. Overall, the simulated data seems to be fairly similar to the observed data.

```{r}
#adding an id so we can run regression on specific simulations
Thousand_Simulated_id <- Thousand_Simulated |>
  mutate(simulation_id = (row_number() - 1) %/% 85 + 1)

# regression x 1000
r_squared_values <- map_dbl(1:1000, function(i) {
  simulated_subset <- Thousand_Simulated_id |>
    filter(simulation_id == i)  

  merged_data <- inner_join(summary_data, simulated_subset, by = "avg_soldiers")

  sub_model <- lm(log(avg_murder_rate) ~ simulated_murder_rate, data = merged_data)

  summary(sub_model)$r.squared
})

```

```{r}
r_squared_data <- data.frame(r_squared_values = r_squared_values)

ggplot(r_squared_data, aes(x = r_squared_values)) +
  geom_histogram(binwidth = 0.005, fill = "darkred", alpha = 0.5, color = "black") +
  geom_vline(xintercept = 0.0129, color = "blue", linetype = "dotted", size = 1) +
  labs(title = "Simulated R^2",
       x = "R^2",
       y = "Frequency") +
  theme_bw()
```

In the plot above, it shows the distribution of simulated R² values from the model. The blue dotted line represents the observed R² from the actual data. Most of the simulated R² values are clustered near zero, showing that the model explains very little of the variation. Overall, the observed R² is slightly higher than most of the simulated values, but the model still does not do a good job predicting murder rates. This suggests that military personnel alone is not a strong predictor of murder rates and other factors are likely more important.
